Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-17 cordillera-cycle.tex
Add warm-base cover fraction map.
Re-read intro to half-discussion and add notes.

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,kyr) and~2
  (29.5--16.9\,kyr). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional ocean sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,kyr).

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. At the last glacial maximum, a
contiguous blanket of ice extended from the Alaska Range in the north to the
North Cascades in the south \citep[Fig.~\ref{fig:locmap};][Plate XIV]
{Chamberlin.1894}. This ice mass is known as the former Cordilleran ice sheet
\citep{Dawson.1888}.

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover
\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
\citealp{Dyke.2004}), directions of past sliding velocities
(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
of former
melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
{Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
and the offshore sedimental record \citep{Cosma.etal.2008, Davies.etal.2011}.

@@

The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle. Although numerical modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet more than twenty years
ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991}), the
ubiquitously mountainous
topography of the region has presented a major challenge to its application.
In fact, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep[e.g.,][]{Bueler.Brown.2009} allows
high-resolution numerical modelling of palaeo-glacier flow on mountainous
terrain over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. In the
North
American Cordillera, the topographic complexity induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2013}.

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model \citep{PISM-authors.2014}. The model inputs basal topography,
sea level, geothermal
heat flux and climate forcing, and computes the evolution of ice extent
and thickness in time, the thermal and dynamic state of the ice sheet, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

To reconstruct the successive phases of growth and decay of the last Cordilleran
ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
by applying time-dependent temperature offsets derived from multiple
palaeo-temperature proxy records. Each simulation starts from assumed ice-free
conditions at -120\,kyr, and runs to present. These computations were
performed on 16 to 128 cores at the Swedish National Supercomputing
Center.

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep[Eqn.~15]{Winkelmann.etal.2011}.
Ice softness depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air
temperature from the atmospheric forcing provides the upper boundary condition
to the ice enthalpy model. Temperature is computed subglacially to a~depth of
3\,km, where it is conditioned by a~geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{Blackwell.Richards.2004}.

A~pseudo-plastic sliding law \citep{PISM-authors.2014} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments.
The till effective pressure is related to the modelled amount of water in the
till, using a formula derived from laboratory experiments with till extracted
from an Antarctic ice stream \citep[Eqn.~2]{Tulaczyk.etal.2000}. Basal
topography is
derived from the ETOPO1 combined topography and bathymetry dataset with
a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.

@@

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, monthly precipitation, and monthly standard
deviation of near-surface temperature by a~temperature-index (positive
degree-day) model \citep[e.g.,]{Hock.2003}. Ice accumulation is equal to
precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep[Eqn.~6]{Calov.Greve.2005} is
numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, after excluding variability
associated with the seasonal cycle itself \citep[cf.][]
{Seguinot.Rogozhina.2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006} over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study \citep{Seguinot.etal.2013}.

The lapse-rate correction ${\Delta}T_{LR}$ is computed using the NARR surface
geopotential height a reference topography $z_{ref}$,
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [z_{s}(t, x, y)-z_{ref}] \\
                            &= -\gamma [h(t, x, y)+z_{b}(t, x, y)-z_{ref}],
\end{align}

@@

For all proxy records, we firstly run a low-resolution simulation, using
a~10\,km, 300 by 150~points horizontal grid for SSA and SIA velocities,
up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
For the GRIP and EPICA records, we then run a second simulation using a higher
horizontal resolution of 6\,km, up to 101~ice layers, and 61~bedrock
layers. In the rest of this paper, we refer to different model runs by the
proxy record used for palaeo-temperature forcing.

@@

Despite large differences in input temperature offset time-series, the model
output presents consistent features that can be observed across the range of
forcing data used. In all simulations, total ice volume is relatively low
during most of the glacial cycle, while two major glaciation events occur
during Marine Isotope Stage (MIS)~4 (61.9--55.4\,kyr) and MIS~2
(29.5--16.9\,kyr, Fig.~\ref{fig:timeseries}). A local ice minimum is
consistently attained during MIS~3 (52.2--45.6\,kyr). However, the
magnitude and timing of these events depend significantly on the choice of
proxy record used for palaeo-temperature forcing.

Simulations forced by the Greenland ice core records (GRIP, NGRIP) produce
highest variability in modelled ice volume throughout the glaciation history.
Model runs using oceanic and Antarctic records generally result in lower
modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus
suggesting lower ice volume variability throughout the simulation length. The
NGRIP palaeo-temperature forcing is the only one that results in larger
glaciation during MIS~4 than during the last glacial maximum.

While simulations driven by the Antarctic (EPICA, Vostok) and GRIP records
reach a last glacial maximum in ice volume between 19.5 and 16.9\,kyr,
those steered by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it
several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
deglaciation prior to 20\,kyr due to a local temperature maximum that is
inconsistent with all other records used. The ODP~1020 simulation predicts an
early last glacial maximum in ice volume at 29.5\,kyr, followed by
slower deglaciation than modelled using other records. Finally, model runs
informed by Antarctic ice core records produce a continuous, rapid deglaciation
after the last glacial maximum, whereas the simulations driven by the GRIP
record produce a rapid deglaciation in three steps, including stand-stills and
readvances of the ice margin.

Despite the different timing, snapshots of model output show relatively
consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
different scaling factors applied to palaeo-temperature proxy records, the
ice sheet geometry at the last glacial maximum appears very similar from one
simulation to the next. It includes a central divide located at about
3\,500\,m elevation along the spine of the Rocky Mountains. In all
simulations, a central ice cap persists over the Skeena mountains during MIS~3
between the two glaciation events. However, the size of this ice cap, as well
as the magnitude of the MIS~4 glaciation, depends sensitively on the choice of
palaeo-temperature proxy record used to drive the model.

@@

@@

Among simulations presented here, only those forced by the GRIP, EPICA and
Vostok records place the last glacial maximum within this range
(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
that pre-dates reconstructed ages by several thousand years. Concerning the
simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
deglaciation is caused by an early warming sealed in the palaeo-temperature records
(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
early warming is a local effect, corresponding to a weakening of the California
current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
waters southwards
along the western coast of North America, is thought to have weakened in the
presence of the Cordilleran ice sheet, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
during the last glacial maximum \citep{Herbert.etal.2001}.

@@

These results make us believe that a palaeo-temperature proxy record located
in the direct vicinity of the former Cordilleran ice sheet would be of great
help to understand its glacial history. Because, to our knowledge, no such
record yet exists, we focus the rest of our analysis on simulations forced by
temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores.

@@

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
Mountains exists for about 29\,kyr in total over MIS~2 and~4.

With the exception of the oceanic front and the northern foothills of the Alaska
Range, most of the maximal extent of the ice sheet is associated with short
durations of ice cover, generally a few thousand years, distributed over one
(GRIP) or two (EPICA) glaciations. This shows that, in our simulations, the
most extensive stages of the Cordilleran ice sheet are both short-lived and
out of balance with contemporary climate states, which confirms
previous interpretations of the geological record \citep{Clague.etal.1980,
Stroeven.etal.2010}.

However, our simulations depict, independently of the choice of record used for
palaeo-temperature forcing, a central Skeena mountain ice cap persisting
throughout the entire last glacial cycle (Fig.~\ref{fig:snapshots} and
\ref{fig:duration}). In particular, this ice cap persists between the MIS~4
and MIS~2 glaciations, providing a nucleation centre for the onset of glacial
readvance towards the last glacial maximum.

@@

In addition, the erosional landscape of the Skeena mountain bears one of the
strongest glacial imprint (?) found within the area formerly covered by the
Cordilleran ice sheet \aref. Here we suggest that persistent ice cover
associated with basal ice temperatures at the pressure-melting point
(Fig.~\ref{fig:warmbase}) may partly explain this spectacular landscape.

@@

% fig:warmbase
\begin{figure*}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,kyr. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.
           \note{indicate location of Skeena Mountains.}}
  \label{fig:warmbase}
\end{figure*}

>>>>

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,kyr) and~2
  (29.5--16.9\,kyr). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,kyr).

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. At the last glacial maximum, a
contiguous blanket of ice extended from the Alaska Range in the north to the
North Cascades in the south (Fig.~\ref{fig:locmap}). This ice mass is known as
the former Cordilleran ice sheet \citep{Dawson.1888}.

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover
\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
\citealp{Dyke.2004}), directions of past sliding velocities
(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
of former
melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
{Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}.

\note{Is the set of references in the above paragraph representative of the
      current state of the field? Am I missing something important here?}

@@

The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle. Although numerical modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet more than twenty years
ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991}), the
ubiquitously mountainous
topography of the region has presented a major challenge to its application.
In fact, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
over millenial time scales \citep{Golledge.etal.2012}. In the North
American Cordillera, this complex topography induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2013}.

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f and stable version~0.6.1), an open-source,
finite-difference, shallow ice sheet model \citep{PISM-authors.2014}.
The model inputs basal topography, sea level, geothermal
heat flux and climate forcing, and computes the evolution of ice extent
and thickness in time, the thermal and dynamic state of the ice sheet, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

To reconstruct the successive phases of growth and decay of the last Cordilleran
ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
by applying time-dependent temperature offsets derived from multiple
palaeo-temperature proxy records. Each simulation starts from assumed ice-free
conditions at -120\,kyr, and runs to present. These computations were
performed on 16 to 128 cores at the Swedish National Supercomputing Centre.

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) by adding both velocity solutions
\citep[Eqn.~15]{Winkelmann.etal.2011}.
Ice softness depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air
temperature from the atmospheric forcing provides the upper boundary condition
to the ice enthalpy model. Temperature is computed subglacially to a~depth of
3\,km, where it is conditioned by a~geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{Blackwell.Richards.2004}.

A~pseudo-plastic sliding law \citep{PISM-authors.2014} \note{ask Ed Bueler if
this will be published somewhere soon} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments.
The till effective pressure is related to the modelled amount of water in the
till, using a formula derived from laboratory experiments with till extracted
from an Antarctic ice stream \citep[Eqn.~2]{Tulaczyk.etal.2000}. Basal
topography is
derived from the ETOPO1 combined topography and bathymetry dataset with
a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level is lowered as
a function of time based on the Spectral Mapping Project (SPECMAP)
\citep[SPECMAP,][]{Imbrie.etal.1989} time scale.
Basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.

@@

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, monthly precipitation, and monthly standard
deviation of near-surface temperature by a~temperature-index model
\citep[e.g.,][]{Hock.2003}. Ice accumulation is equal to precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep[Eqn.~6]{Calov.Greve.2005} is
numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, in order to account for the
effects of spatial and seasonal variations of temperature variability
\citep[Fig.~\ref{fig:atm};][]{Seguinot.2013}, after excluding variability
associated with the seasonal cycle itself \citep[cf.][]
{Seguinot.Rogozhina.2014}. The ablation model incorporates degree-day factors
of 3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006} over the period 1979--2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability
(Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2013}.

The lapse-rate correction ${\Delta}T_{LR}$ is computed using the NARR surface
geopotential height invariant field as a reference topography $z_{ref}$,
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [z_{s}(t, x, y)-z_{ref}] \\
                            &= -\gamma [h(t, x, y)+z_{b}(t, x, y)-z_{ref}],
\end{align}

@@

For all proxy records, we firstly run a low-resolution simulation, using
a~10\,km, 300 by 150~points horizontal grid for SSA and SIA velocities,
up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
For the GRIP and EPICA records, we then run a second simulation using a higher
horizontal resolution of 6\,km, up to 101~ice layers, and 61~bedrock
layers. In the rest of this paper, we occasionally refer to different model
through by the proxy record used for palaeo-temperature forcing.

@@

Despite large differences in input temperature offset time-series, the model
output presents consistent features that can be observed across the range of
forcing data used. In all simulations, total ice volume is relatively low
during most of the glacial cycle, while two major glaciation events occur
during Marine Isotope Stage (MIS)~4 (61.9--55.4\,kyr) and MIS~2
(29.5--16.9\,kyr, Fig.~\ref{fig:timeseries}). A local ice minimum is
consistently attained during MIS~3 (52.2--45.6\,kyr). However, the
magnitude and timing of these three events depend significantly on the choice
of proxy record used for palaeo-temperature forcing.

Simulations forced by the Greenland ice core records (GRIP, NGRIP) produce
highest variability in modelled ice volume throughout the glaciation history.
Model runs using oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
records generally result in lower
modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus
suggesting lower ice volume variability throughout the simulation length. The
NGRIP palaeo-temperature forcing is the only one that results in larger
glaciation during MIS~4 than during the last glacial maximum (MIS~2).

While simulations driven by the Antarctic and GRIP records
reach a last glacial maximum in ice volume between 19.5 and 16.9\,kyr,
those steered by oceanic and the NGRIP records attain it
several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
deglaciation prior to 20\,kyr due to a local temperature maximum that is
inconsistent with all other records used. The ODP~1020 simulation predicts an
early last glacial maximum in ice volume at 29.5\,kyr, followed by
slower deglaciation than modelled using other records. Finally, model runs
informed by Antarctic ice core records produce a continuous, rapid deglaciation
after the last glacial maximum, whereas the simulations driven by the GRIP
record produce a rapid deglaciation in three steps, including stand-stills and
readvances of the ice margin.

Despite the different timing, snapshots of model output show relatively
consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
different scaling factors applied to palaeo-temperature proxy records,
modelled ice sheet geometries at the last glacial maximum appear very similar
from one simulation to the next. They include a central divide located at about
3\,500\,m elevation along the spine of the Rocky Mountains. In all
simulations, a central ice cap persists over the Skeena mountains during MIS~3
between the two glaciation events. However, the size of this ice cap, as well
as the magnitude of the MIS~4 glaciation, depend sensitively on the choice of
palaeo-temperature proxy record used to drive the model.

@@

\note{Could you help me to fill the missing refs in this last sentence?}

@@

Among simulations presented here, only those forced by the GRIP, EPICA and
Vostok records place the last glacial maximum within this range
(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
that pre-dates reconstructed ages by several thousand years. Concerning the
simulations driven by oceanic records, this early deglaciation is caused by an
early warming sealed in the palaeo-temperature reconstructions
(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
early warming is a local effect, corresponding to a weakening of the California
current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
waters southwards
along the western coast of North America, is thought to have weakened in the
presence of the Cordilleran ice sheet, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
during the last glacial maximum \citep{Herbert.etal.2001}.

@@

These results make us believe that a palaeo-temperature proxy record located
in the direct vicinity of the former Cordilleran ice sheet would be of great
help to further understand its glacial history. Because, to our knowledge, no
such record yet exists, we focus the rest of our analysis on simulations forced
by temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores
(Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

@@

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
Mountains exists for about 29\,kyr in total over MIS~2 and~4.
With the exception of the oceanic front and the northern foothills of the Alaska
Range, most of the maximal extent of the ice sheet is associated with short
durations of ice cover, generally a few thousand years, distributed over one
(GRIP) or two (EPICA) glaciations (Fig.~\ref{fig:duration}). This illustrates
that the most extensive stages of the modelled ice sheet are both short-lived
and out of balance with contemporary climate states, which confirms previous
inferences of a short-lived Cordilleran ice sheet based on the geological
record \citep{Clague.etal.1980, Stroeven.etal.2010}.

However, our simulations depict, independently of the choice of record used for
palaeo-temperature forcing, a central Skeena mountain ice cap persisting
throughout the entire last glacial cycle (Fig.~\ref{fig:snapshots} and
\ref{fig:duration}). In particular, this ice cap persists during MIS~3,
providing a nucleation centre for the onset of glacial
readvance towards the last glacial maximum (MIS~2).

@@

In addition, the erosional landscape of the Skeena mountain bears one of the
strongest glacial imprint (?) found within the area formerly covered by the
Cordilleran ice sheet \aref. Here we suggest that persistent ice cover
associated with basal ice temperatures at the pressure-melting point
(Fig.~\ref{fig:warmbase} and ~\ref{fig:warmbase}) may partly explain this
spectacular landscape.

\note{Can we use a reference here to support the point of strong glacial
      imprint in the Skeena Mountains?}

@@

% fig:warmbase
\begin{figure*}
  \includegraphics{warmbase}
  \caption{Modelled fraction of warm-based ice cover during the ice-covered
           period. Note the dominance of warm-based conditions on the
           continental shelf and major glacial troughs of the coastal ranges.
           Hatches indicate areas that were covered by cold ice only.}
  \label{fig:warmbase}
\end{figure*}

% fig:warmfrac
\begin{figure*}
  \includegraphics{warmfrac}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,kyr. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.
           \note{indicate location of Skeena Mountains.}}
  \label{fig:warmfrac}
\end{figure*}
